A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-02-16 papers/execution-model/model.tex
sec 3

The execution model is an abstraction of the Tezos blockchain \cite{Tezos}
augmented with functional (OCaml) programs that interact with the
blockchain. As the blockchain is realized by a peer-to-peer network of
independent nodes, interaction happens through
\emph{local nodes} that receive requests to originate and invoke
contracts from programs that run on a particular node. We model the blockchain
itself as a separate, abstract global entity that represents the
current consensual state of the system. Our model does not express
low-level details, but relies on nondeterminism to describe
the possible behaviors of the system.

@@

\begin{definition}%[Account]
  An implicit \emph{account} is represented by a key pair $\langle
  %\ALS,
  \PAK, \PUK
  %, \PKH
  \rangle $, where
  $\PAK$ is its private key and $\PUK$ is its public key.
\end{definition}

\begin{definition}%[Operation]
  \emph{Operations} are defined by the following grammar:
  \begin{align*}
    \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
    \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
  \end{align*}
  where \NTEZ\ and \MTEZ\ are tokens, \ADDR\ is either a public
  key for an implicit account  or a public hash for a smart contract,
  \PARAMETER\ is an argument passed to the smart contract's script,
  which is empty if it is a transfer to an implicit account, \CODE\ is
  a script of the smart contract, \INIT\ is an initial value of the
  contract's storage. 
\end{definition}

\begin{definition}%[Query]
\emph{Queries} are defined by the following grammar:
\begin{align*}
  \QRY & ::=
  \GETCOUNTER\PUK  
  \\&\mid \GETBALANCE\ADDR
  \\&\mid \GETSTATUS\OPH
  \\&\mid \GETSTORAGE\PUH
  % \\&\mid \GETCODE\PUH 
  % \\&\mid \GETTYPE\PUH
  \\&\mid \GETCONTRACT\OPH
\end{align*}
\end{definition}

\begin{definition}%[Program]
\emph{Expressions} are defined by the following grammar:
\begin{align*}
  \CONSTANT & ::= \INT %\mid \ADDR
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR \mid (\EXPR,\EXPR) \mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
	\\&\mid \EXPR\ \PLUS\ \EXPR\ \mid \EXPR\ \MINUS\ \EXPR\ \mid \EXPR\ \EQUAL\ \EXPR\ \mid \EXPR\ \LESS\ \EXPR
	\\&\mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\&\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR \mid \CONTRACT\ \TYPE\ \EXPR \mid \TOSTRING\
  \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter \mid type
\end{align*}
We let $\EXPRS$ range over sets of expressions.

We let $\ECS$ range over evaluation contexts. They are defined as
follows.
\begin{align*}
  \EC{} & :: = \SC[\overline\VAL\ \EC{}\ \overline\EXPR] \mid \RAISE\ \EC{} \mid \TRY\ \EC{}\ \EXCEPT\ \EXPR
  \mid \MATCH\ \EC{}\ \WITH\ \PATTERN\to\EXPR \dots
\end{align*}
Here $\SC$ ranges over the remaining syntactic constructors, which are
treated uniformly: evaluation proceeds from left to right.
\begin{align*}
  \VAL & ::= \CONSTANT \mid \STATUS \mid \ERROR \mid \lambda x.\EXPR \mid(\VAL, \VAL) \mid
         \NIL \mid \CONS\ \VAL\ \VAL \mid \LEFT\ \VAL \mid \RIGHT\
         \VAL \mid \SOME\ \VAL\mid \NONE
\end{align*}
\end{definition}

Example pseudo code
\begin{align*}
  \OPH &=  \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR \\
       & \MATCH\ {status\ \OPH}\ \WITH \\
       & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
       & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
         blocks confirmation}\\
       & \mid \STATUSTIMEOUT \to \textrm{lost, try again}
\end{align*}
Grammar of types
\begin{align*}
  \TYPE, \TYPEU & ::= \TADDR \mid \TOPH \mid
                  \TCONTRACT\ \TYPE \mid
                  % \TPUH\ \TYPE \mid
                  \TPUK \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
          \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TSTATUS \\
  & \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE \mid \TEXCEPTION
\end{align*}

>>>>

The execution model is an abstraction of the Tezos blockchain \cite{Tezos}
augmented with functional (OCaml) programs that interact with the
blockchain. As the blockchain is realized by a peer-to-peer network of
independent nodes, interaction happens through
\emph{local nodes} that receive requests to originate and invoke
contracts from programs that run on a particular node. We model the blockchain
itself as a separate, abstract global entity that represents the
current consensual state of the system. Our model does not express
low-level details, but relies on nondeterminism to describe
the possible behaviors of the system. In particular, we do
\textbf{not} formalize the execution of the smart contracts
themselves, we rather consider them as black boxes that execute in a
safe-safe way. Tezos' smart contract language Michelson and its
properties have been formalized elsewhere \cite{type-safe-michelson}.

@@

A local node runs on behalf of authorities, which are called
\emph{accounts} in Tezos. An account is represented by a key pair $\Angle{
  \PAK, \PUK
} $, where $\PAK$ is a private key and $\PUK$ the corresponding public
key in a public key encryption scheme.

The local node offers operations to transfer tokens from one account
to another, to invoke a contract, and to originate a contract on the blockchain.
\begin{align*}
  \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
  \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
\end{align*}
Here, \NTEZ\ and \MTEZ\ are tokens, \PUK\ is the public key of the
sender, \ADDR\ is either a public 
key for an implicit account (in case of a simple transfer)  or a
public hash for a smart contract (for an invocation), \PARAMETER\ is
the argument passed to the smart contract, 
which is empty for a simple transfer. Furthermore,  \CODE\ is
a script of a smart contract and \INIT\ is the initial value of the
contract's storage. Each operation returns an \emph{operation hash}
\OPH, on which we can query the status of the operation.

The local node also offer several ways to query the current state of
the blockchain.
\emph{Queries} are defined by the following grammar:
\begin{align*}
  \QRY & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE\ADDR
  \mid \GETSTATUS\OPH
  \mid \GETSTORAGE\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
  \mid \GETCONTRACT\OPH
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the code of a contract by the
operation hash of its originating transaction.

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR \mid
  \CONTRACT\ \TYPE\ \EXPR \mid \TOSTRING\
  \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
  &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter
         \mid type
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TADDR \mid \TOPH \mid
                  \TCONTRACT\ \TYPE \mid
                  % \TPUH\ \TYPE \mid
                  \TPUK \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
          \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by closed expressions that run on the
local node. Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types and exceptions. Moreover, there are
primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, the three constructs 
\CONTRACT\ \TYPE\ \EXPR, \TOSTRING\  \EXPR, and \FROMSTRING\ \TYPE\
\EXPR\ are casts between untyped and typed representations of
contracts and their parameters.

\begin{figure}[tp]
\begin{align*}
  \ECN{} & :: = \ECHOLE \mid \SC[\overline\VAL\ \ECN{}\ \overline\EXPR] \mid \RAISE\ \ECN{} \mid \TRY\ \ECN{}\ \EXCEPT\ \EXPR
  \mid \MATCH\ \ECN{}\ \WITH\ \PATTERN\to\EXPR \dots
  \\
  \VAL & ::= \CONSTANT \mid \STATUS \mid \ERROR \mid \lambda x.\EXPR \mid(\VAL, \VAL) \mid
         \NIL \mid \CONS\ \VAL\ \VAL \mid \LEFT\ \VAL \mid \RIGHT\
         \VAL \mid \SOME\ \VAL\mid \NONE
\end{align*}
  \caption{Evaluation contexts and values}
  \label{fig:evaluation-contexts-values}
\end{figure}
Figure~\ref{fig:evaluation-contexts-values} defines evaluation
contexts $\ECS$ and values $\VAL$.
Here $\SC$ ranges over the remaining syntactic constructors, which are
treated uniformly: evaluation proceeds from left to right. Values are
standard for call-by-value lambda calculus.

\begin{figure}[tp]
  \begin{align*}
    \OPH &=  \TRANSFER[\STRING]\NTEZ\PUK\PUH\MTEZ \\
    \\
         & \MATCH\ {status\ \OPH}\ \WITH \\
         & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
         & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
           blocks confirmation}\\
         & \mid \STATUSTIMEOUT \to \textrm{lost, try again}
  \end{align*}
  \caption{Pseudocode example}
  \label{fig:pseudocode-example}
\end{figure}
Figure~\ref{fig:pseudocode-example} shows an expression that invokes a
contract at \PUH\ with argument \STRING.

Grammar of types
